Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-19 analysis/paper/paper.Rmd
Writing in progress - this version contains road network projection details in appendix A for reference. The details will reduced and moved into the main text in future versions.

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales. More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [e.g. @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package. We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia. The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose. Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].

@@

Continue reviewing pdfs in downloads -

-->
```
<!-- One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). -->

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)). This allows projection of the forest road network from historic forest harvest data. Approximately 55% of the area is non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010], so the area provides opportunities to observe the responses of road projection algorithms to barriers. Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife species that are sensitive to roads including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple model informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (https://osf.io/8vmqh/). We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.

<!-- TO DO: add link to OSF repository once public-->

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock are determined independently in the simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023].

@@

To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott_roads_2023]. The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).

@@

In order to assess the performance of the projections, five characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, measured as whether or not roads are present in each pixel; 2) Road density, measured as density of roads in meters per hectare (m/ha); 3) Distance to nearest road, measured as distance from each raster cell to the nearest road; 4) Road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and 5) Forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020], and road density is often used in studies of anthropogenic impacts on wildlife [@beazley_road_2004; @apps_factors_2006; more]. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package. All road characteristics were stored in raster format.

@@

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the timber supply area and within cutblocks by masking the road network characteristic rasters with cutblock polygons and calculating means within cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: 1) pre-existing footprint; 2) agree roadless, meaning that both projected and observed show no footprint; 3) agree roaded, meaning that both projected and observed show new footprint; 4) false positive, meaning the projection shows a footprint but no footprint was observed; and 5) false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). Within all mapped cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks. These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain (Figure \@ref(fig:proj-meth-fig)). The Hardy method projects more tortuous roads within cutblocks, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

Another fundamental limitation of all these methods is that they generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy. Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.

@@

@@

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy's Landis-II extension [-@hardy_forest_2019; -@hardy_landis-ii_2023] can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provides a comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model. Thus, we were not surprised to less accurate projections of road density and presence than Hardy found in less mountainous regions of eastern Canada with a model that more carefully accounted for water crossing costs [-@hardy_landis-ii_2023]. The stream crossing behaviour of road projection algorithms in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains. A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. The performance of these algorithms varies with terrain; in topographically complex landscape high resolution cost surfaces that account for the costs of stream and river crossings will address some but not all of the challenges of forest road projection in mountain regions.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. The lack of consistent accurate information on the location of resource roads across Canada and elsewhere [@poley_identifying_2022](Poley, ) Better harvest & existing road data would help errors in projections from errors in the input data (cite lit in coops paper)

>>>>

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, very few of these projections have included road networks. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales. More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package. We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia. The adequacy of a road network projection methods depends on the purpose of the projection, so it is important to select performance metrics that reflect the purpose. Given our interest in projecting impacts on wildlife we focus on several common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023; @stewart_climate-informed_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].

@@

-->
```

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because it is a mountainous areas with approximately 55% of non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010]. Thus, the area provides opportunities to observe the responses of road projection algorithms to steep complex terrain. Forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)), which allows projection of the forest road network from historic forest harvest data.  Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

Road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

We developed a simple model of road building cost informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>. We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>. Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.

<!-- TO DO: number analysis scripts to clarify workflow -->

@@

We focus on methods that link multiple target locations to an existing road network for strategic purposes, described by Dean (1997) as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network. We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail; in this manuscript we focus on comparing the dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAP). For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design. The DLCP approach solves each STAP sequentially, starting with target that is nearest to an existing road. The existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target. The method was first developed by Anderson and Nelson (2004) who noted that the projected network is sensitive to the ordering of the STAP and recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network. These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node in graph with a tree that minimizes edge weights and does not include loops. This method differs from the minimum spanning tree method used by Clark et al. (2000) who used Euclidean distances rather than least costs for edge weights.

@@

To help make these road network projection methods more easily and widely useable, we implemented the DLCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).

@@

To assess performance we calculated five characteristics of the projected and observed road networks: road presence, measured as whether or not roads are present in each pixel; road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011]. The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g. @labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.

TO DO: Sarah are these statements still correct? Please revise as necessary.

@@

Aspatial performance of the projections was assessed by comparing projected and observed mean values of each road network characteristic across the timber supply area and within cutblocks. We further distinguished mapped cutblocks that contain observed roads from those that do not to reduce errors associated with open areas that have been misidentified as cutblocks. Spatial performance was assessed in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Second, performance according to the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed by categorizing pixels of a projected raster into 5 categories compared with the observed raster: pre-existing footprint; agree roadless, meaning that both projected and observed show no footprint; agree roaded, meaning that both projected and observed show new footprint; false positive, meaning the projection shows a footprint but no footprint was observed; and false negative, meaning the projection shows no footprint but a footpint was observed. These categories were summarized for each projection and characteristic and used to calculate sensitivity, precision and F-measure of each metric [@lee_prediction_2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). Within all mapped cutblocks (Figure \@(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@(fig:aspat-perf-fig-all)). These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuousity within cutblocks (Figure \@ref(fig:proj-meth-fig)). None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

Another fundamental limitation of these methods is that they generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy. Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" scenario where cutblocks were the only disturbance considered.

@@

TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.

@@

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide a complementary and comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model. Thus, we were not surprised to see less accurate projections of road density and presence than in less mountainous regions of eastern Canada with more careful accounting for the cost of water crossings [@hardy_landis-ii_2023]. Water crossing behaviour in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains. A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density becomes less accurate as raster cell resolution increases.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022] that users should keep in mind interpreting discrepancies between observed and projected road networks.  
Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are more costly to develop and use. Before investing in additional complexity, it is important to recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor [e.g. @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020]. In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.
